---
title: "final_project"
output: html_document
author: "Simran Kota, Yusef Haswarey"
---

# Packages Used
```{r setup, include=TRUE}
library(tidyverse)
library(randomForest)
library(caret)
library(ggpubr)
```

# Research Question

We are attempting to explore the relationship between various features of a house (ex. number of rooms, square footage, year built, etc.) on the future value of the house.

# Data Source and Definitions

The [USA Housing Dataset](https://www.kaggle.com/gpandi007/usa-housing-dataset?select=housing_train.csv) from Kaggle provides data regarding sales prices with respect to various houses in the US. It has 81 features describing each record for 1460 records.

``` {r}
dat <- read.csv("housing_data.csv")
str(dat)
```

We don't care about the Id column, so we will drop it.
``` {r}
dat <- dat[,-c(1)]
str(dat)
```


While the full variable descriptions can be found in the data_description.txt file in the repository, we have included a summary below:


MSSubClass: type of house, categorical

MSZoning: zone of house, categorical
	
LotFrontage: Linear feet of street connected to house, continuous

LotArea: Lot size in square feet, continuous

Street: Type of road access to house, categorical
       	
Alley: Type of alley access to house, categorical
		
LotShape: General shape of property, categorical
       
LandContour: Flatness of the property, categorical
		
Utilities: Type of utilities available, categorical
	
LotConfig: Type of lot, categorical
	
LandSlope: Slope of property, categorical
	
Neighborhood: Physical location, categorical
			
Condition1: Proximity to various conditions, categorical
	
Condition2: Proximity to various conditions (if more than one is present), categorical
	
BldgType: Type of home, categorical
	
HouseStyle: Style of home, categorical
	
OverallQual: Rating of the overall material and finish of the house, categorical
	
OverallCond: Rating of the overall condition of the house, categorical
		
YearBuilt: Original construction date, continuous

YearRemodAdd: Remodel date (same as construction date if no remodeling or additions), continuous

RoofStyle: Type of roof, categorical
		
RoofMatl: Roof material, categorical
		
Exterior1st: Exterior covering on house, categorical
	
Exterior2nd: Exterior covering on house (if more than one material), categorical
	
MasVnrType: Masonry veneer type, categorical
	
MasVnrArea: Masonry veneer area in square feet, categorical

ExterQual: Evaluates the quality of the material on the exterior, categorical
		
ExterCond: Evaluates the present condition of the material on the exterior, categorical
		
Foundation: Type of foundation, categorical
		
BsmtQual: Evaluates the height of the basement, categorical
		
BsmtCond: Evaluates the general condition of the basement, categorical
	
BsmtExposure: Refers to walkout or garden level walls, categorical
	
BsmtFinType1: Rating of basement finished area, categorical
		
BsmtFinSF1: Type 1 finished square feet, continuous

BsmtFinType2: Rating of basement finished area (if multiple types), categorical

BsmtFinSF2: Type 2 finished square feet, continuous

BsmtUnfSF: Unfinished square feet of basement area, continuous

TotalBsmtSF: Total square feet of basement area, continuous

Heating: Type of heating, categorical
		
HeatingQC: Heating quality and condition, categorical
		
CentralAir: Central air conditioning, categorical (binary)
		
Electrical: Electrical system, categorical
		
1stFlrSF: First Floor square feet, continuous
 
2ndFlrSF: Second floor square feet, continuous

LowQualFinSF: Low quality finished square feet (all floors), continuous

GrLivArea: Above ground living area square feet, continuous

BsmtFullBath: Basement full bathrooms, continuous

BsmtHalfBath: Basement half bathrooms, continuous

FullBath: Full bathrooms above ground, continuous

HalfBath: Half baths above ground, continuous

Bedroom: Bedrooms above ground (does NOT include basement bedrooms), continuous

Kitchen: Kitchens above grade, continuous

KitchenQual: Kitchen quality, categorical
       	
TotRmsAbvGrd: Total rooms above ground (does not include bathrooms), continuous

Functional: Home functionality, categorical
		
Fireplaces: Number of fireplaces, continuous

FireplaceQu: Fireplace quality, categorical
		
GarageType: Garage location, categorical
		
GarageYrBlt: Year garage was built, continuous
		
GarageFinish: Interior finish of the garage, categorical
		
GarageCars: Size of garage in car capacity, continuous

GarageArea: Size of garage in square feet, continuous

GarageQual: Garage quality, categorical
		
GarageCond: Garage condition, categorical
		
PavedDrive: Paved driveway, categorical
		
WoodDeckSF: Wood deck area in square feet, continuous

OpenPorchSF: Open porch area in square feet, continuous

EnclosedPorch: Enclosed porch area in square feet, continuous

3SsnPorch: Three season porch area in square feet, continuous

ScreenPorch: Screen porch area in square feet, continuous

PoolArea: Pool area in square feet, continuous

PoolQC: Pool quality, categorical
		
Fence: Fence quality, categorical
	
MiscFeature: Miscellaneous feature not covered in other categories, categorical
		
MiscVal: Value of miscellaneous feature, continuous

MoSold: Month Sold (MM), continuous

YrSold: Year Sold (YYYY), continuous

SaleType: Type of sale, categorical
		
SaleCondition: Condition of sale, categorical

SalePrice: Price the house was sold for, continuous

Although some of the variables were automatically mistyped by R, we don't need all the variables for our analysis. We will first drop some variables to reduce the dataset to the data required, then retype the variables as needed.

``` {r}
dat <- subset(dat, select=-c(Condition1, Condition2, Exterior2nd, MasVnrType, MasVnrArea, BsmtExposure, BsmtFinType2, BsmtFinSF2, SaleType, SaleCondition, GarageYrBlt))
dim(dat)
```

We are left with 69 features out of the original 81. We still have 1460 records, however, it is highly likely at least some of these have null values. We will examine these and clean them up next.

``` {r}
colSums(is.na(dat))
```

``` {r}
dat$LotFrontage[is.na(dat$LotFrontage)] <- 0
dat$Alley[is.na(dat$Alley)] <- "None"
dat$BsmtQual[is.na(dat$BsmtQual)] <- "None"
dat$BsmtCond[is.na(dat$BsmtCond)] <- "None"
dat$BsmtFinType1[is.na(dat$BsmtFinType1)] <- "None"
dat$Electrical[is.na(dat$Electrical)] <- "Unknown"
dat$FireplaceQu[is.na(dat$FireplaceQu)] <- "None"
dat$GarageType[is.na(dat$GarageType)] <- "None"
dat$GarageFinish[is.na(dat$GarageFinish)] <- "None"
dat$GarageQual[is.na(dat$GarageQual)] <- "None"
dat$GarageCond[is.na(dat$GarageCond)] <- "None"
dat$PoolQC[is.na(dat$PoolQC)] <- "None"
dat$Fence[is.na(dat$Fence)] <- "None"
dat$MiscFeature[is.na(dat$MiscFeature)] <- "None"
```

Checking our work:

``` {r}
colSums(is.na(dat))
```

Now that we have removed all the nulls, we will retype our variables.

``` {r}
dat$MSSubClass <- as.factor(dat$MSSubClass)
dat$MSZoning <- as.factor(dat$MSZoning)
dat$Street <- as.factor(dat$Street)
dat$Alley <- as.factor(dat$Alley)
dat$LotShape <- as.factor(dat$LotShape)
dat$LandContour <- as.factor(dat$LandContour)
dat$Utilities <- as.factor(dat$Utilities)
dat$LotConfig <- as.factor(dat$LotConfig)
dat$LandSlope <- as.factor(dat$LandSlope)
dat$Neighborhood <- as.factor(dat$Neighborhood)
dat$BldgType <- as.factor(dat$BldgType)
dat$HouseStyle <- as.factor(dat$HouseStyle)
dat$OverallQual <- as.factor(dat$OverallQual)
dat$OverallCond <- as.factor(dat$OverallCond)
dat$RoofStyle <- as.factor(dat$RoofStyle)
dat$RoofMatl <- as.factor(dat$RoofMatl)
dat$Exterior1st <- as.factor(dat$Exterior1st)
dat$ExterQual <- as.factor(dat$ExterQual)
dat$ExterCond <- as.factor(dat$ExterCond)
dat$Foundation <- as.factor(dat$Foundation)
dat$BsmtQual <- as.factor(dat$BsmtQual)
dat$BsmtCond <- as.factor(dat$BsmtCond)
dat$BsmtFinType1 <- as.factor(dat$BsmtFinType1)
dat$Heating <- as.factor(dat$Heating)
dat$HeatingQC <- as.factor(dat$HeatingQC)
dat$CentralAir <- as.factor(dat$CentralAir)
dat$Electrical <- as.factor(dat$Electrical)
dat$KitchenQual <- as.factor(dat$KitchenQual)
dat$Functional <- as.factor(dat$Functional)
dat$FireplaceQu <- as.factor(dat$FireplaceQu)
dat$GarageType <- as.factor(dat$GarageType)
dat$GarageFinish <- as.factor(dat$GarageFinish)
dat$GarageQual <- as.factor(dat$GarageQual)
dat$GarageCond <- as.factor(dat$GarageCond)
dat$PavedDrive <- as.factor(dat$PavedDrive)
dat$PoolQC <- as.factor(dat$PoolQC)
dat$Fence <- as.factor(dat$Fence)
dat$MiscFeature <- as.factor(dat$MiscFeature)
```

Checking our work:
``` {r}
str(dat)
```
# Exploratory Data Analysis

``` {r}
for (col in c(names(dat))) {
  g <- ggplot(data=dat, aes_string(x=col))
  if (is.numeric(dat[[col]])) {
    g <- g + geom_histogram(bins=30)
  } else {
    g <- g + geom_bar()
  }
  print(g)
}
```


# Primary Method - Random Forests

``` {r}
set.seed(1234567890)

n <- nrow(dat)

tv.split <- sample(rep(0:1, c(round(n*.3),n-2*round(n*.15))),n)

dat.train <- dat[tv.split==1,]

dat.valid <- dat[tv.split==0,]

housing.randomForest <- randomForest(formula = SalePrice ~ ., data = dat)

plot(housing.randomForest)

variable.Importance <- data.frame(housing.randomForest$importance) %>%
  arrange(desc(IncNodePurity)) %>%
  top_n(10)

(ggplot(data = variable.Importance,
       aes(x = reorder(row.names(variable.Importance), IncNodePurity), y = IncNodePurity))
        + geom_col() + coord_flip() + ggtitle("Top 10 important variables"))
```

```{r}
x.valid <- dat.valid[setdiff(names(dat.valid), "SalePrice")]

y.valid <- dat.valid$SalePrice

randomforest.trained <- randomForest(formula = SalePrice ~ ., data = dat.train)

find.RMSPE <- function(true, predicted, df) {
  return(sqrt(sum((predicted - true)^2)/nrow(df)))
}

predictions.valid <- predict(randomforest.trained, dat.valid)

RMSPE <- find.RMSPE(dat.valid$SalePrice, predictions.valid, dat.valid)

RMSPE
```

```{r}
randomforest.validated <- randomForest(formula = SalePrice ~ ., data = dat.train,
                                       xtest = x.valid, ytest = y.valid)

tibble(`Out of Bag Error` = sqrt(randomforest.validated$mse),
       `Validation Error` = sqrt(randomforest.validated$test$mse),
       ntrees = 1:randomforest.validated$ntree) %>%
  gather(Grouping, RMSE, -ntrees) %>%
  ggplot(aes(ntrees, RMSE, color = Grouping)) +
  geom_line() +
  scale_y_continuous(labels = scales::dollar) +
  xlab("Number of trees")
```


